---
title: "Associations"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this report, we will show the different associations performed.

```{r}
library(qqman)
library(kableExtra)
library(tidyverse)
library(GenomicRanges)
library(SNPassoc)
library(epiDisplay)
library(cowplot)
library(VariantAnnotation)
library(matrixStats)
library(readxl)
```

# GWAS

First, we run a GWAS analysis with the imputed SNPs passing the QC. We adjusted all the models for sex and 10 first genetic principal components (PCs). We used the logistic regression as implemented in plink. 


```{r}
lambdaClayton <- function(x, trim = 0.5) {
    xx <- qnorm(1-x)^2
    N <- length(xx)
    obsvd <- sort(xx, na.last=NA)
    expctd <- qchisq(p = (1:N)/(N + 1), 1)
    Nu <- floor(trim * N)
    lambda <- mean(obsvd[1:Nu])/mean(expctd[1:Nu])
    lambda
}
version <- "2021-06-15"
```

```{r, cache = TRUE}
getGWASTab <- function(path){
   tab <- read_table2(path, col_names = FALSE)
colnames(tab) <- c("CHR", "SNP", "BP", "ALLELE", "MODEL", "SAMPS", "BETA", "SE", "P", "REF", "ALT") 
    subset(tab, !is.na(P))
}
single.snp <- getGWASTab(paste0("../results/associations/GWAS/", version, "/Obesity.GWAS.locuszoom.input.gz"))

```

```{r}
qq(single.snp$P)
lambdaClayton(single.snp$P)
```

We do not observe inflation in the GWAS results

```{r}
manhattan(single.snp)
```

We observed `r sum(single.snp$P < 5e-8)` SNPs with genome-wide significant association (P < 5e-8) with the extreme obesity, distributed in chromosomes 1, 6, 7, 12, 16, 17, 18, 19, 22 and 23.

We run two approaches to define loci:

- Clumping with plink: plink generated loci starting by SNPs with genome-wide significance (P < 5.e-8). The loci were extended to include all nearby SNPs with P < 1e-5 and in high LD.
- Depict: depict based on SNPs with P < 1e-5. Performs a similar clustering but also includes nearby genes.

```{r}
lociplink <- read_table2(paste0("../results/associations/GWAS/", version, "/Obesity.GWAS.plink.clumped"),
                     progress = TRUE, col_names = TRUE)
locidepict <- read_delim("../results/associations/GWAS/2021-06-11/assoc_loci.txt",
                     progress = TRUE, col_names = TRUE, comment = "#", delim = "\t")

## Create range with SNPs in SP2
getNum <- function(SP2){
    if (SP2 == "NONE"){
        return(NA)
    }
    a <- strsplit(SP2, ":")[[1]]
    as.numeric(a[seq(2, length(a), 3)])
}
    
lociplink.range <- lociplink %>%
    mutate(minClump = sapply(SP2, function(x) min(getNum(x))),
           maxClump = sapply(SP2, function(x) max(getNum(x))),
           start = pmin(minClump, BP, na.rm = TRUE), 
           end = pmax(minClump, BP, na.rm = TRUE)) 
lociplink.range$id <- paste0("plink", seq_len(nrow(lociplink.range)))
plinkGR <- makeGRangesFromDataFrame(lociplink.range, start.field = "BP", end.field = "BP")
plinkGR$id <- paste0("plink", seq_len(length(plinkGR)))

locidepict$id <- paste0("depict", seq_len(nrow(locidepict)))
depictGR <- makeGRangesFromDataFrame(locidepict)
depictGR$id <- paste0("depict", seq_len(length(depictGR)))
ov <- findOverlaps(plinkGR, depictGR)
plink.un <- lociplink.range[-from(ov), ]
comGR <- c(plinkGR[-from(ov)], depictGR[unique(to(ov))])
```

Overall, we found `r length(plinkGR)` loci with plink and `r length(depictGR)` loci with depict. After combining both sets, and discarding depict loci without genome-wide significant SNPs, we have a total  `r length(comGR)` loci with genome-wide significance.

Next, we removed those loci with an overlap > 30% with segmental duplications. 

```{r}
LCRs <- read.table("/home/SHARED/DATA/REFERENCES/GRCh37/SegmentalDups/RLCRs_chrNCBI_no_Repeat_Masker.txt", comment.char = "", header = FALSE, as.is = TRUE)
LCRGR <- makeGRangesFromDataFrame(LCRs, seqnames.field = "V1",
                                  start.field = "V2", end.field = "V3")

over.LCR <- findOverlaps(comGR, LCRGR)
comGR.filt <- comGR[-unique(from(over.LCR))]
good <- sapply(unique(from(over.LCR)), function(i){
    
    segDup.sel <-  LCRGR[to(ov)[from(ov) == i]]
    sum(width( segDup.sel)) < 0.3*width(comGR[i])
})
comGR.filt <- c(comGR.filt, comGR[unique(from(over.LCR))][good])
```

We got a final set of `r length(comGR.filt)` loci with significant genome-wide association with extreme obesity and outside segmental duplications regions. 


## Imputation artifacts

We evaluated whether the hits might be caused by artifacts in the imputation.

```{r}
getSNPimputationStatus <- function(cohort, selSNPs){
    
    path <- "../results/postImputation/MichiganOutput/"
    chrom <- c(1:22, "X")
    infoL <- lapply(chrom, function(chr){
        a <- read_table2(paste0(path, cohort, "/chr", chr, ".info.gz"))
        subset(a, SNP %in% selSNPs)
    })
    infoDf <- Reduce(rbind, infoL)
    infoDf$Cohort <- cohort
    infoDf
}
info.full.list <- lapply(c("Epicuro", paste0("Obesity", 1:3)), 
                    getSNPimputationStatus, selSNPs = single.snp$SNP)
info.full.df <- Reduce(rbind, info.full.list)
```


```{r}
snp.imp.tab <- info.full.df %>%
  dplyr::select(SNP, Genotyped, Cohort) %>%
  spread(Cohort, Genotyped)

geno.SNPs <- subset(snp.imp.tab, Epicuro == "Genotyped" & 
                      Obesity1 == "Genotyped" & 
                      Obesity2 == "Genotyped" &
                      Obesity3 == "Genotyped" )$SNP
geno.SNPs2 <- subset(snp.imp.tab, Epicuro == "Genotyped" & 
                      Obesity3 == "Genotyped" )$SNP
```


```{r}
manhattan(subset(single.snp, SNP %in% geno.SNPs))

```
When restricting the GWAS to SNPs genotypes in the 4 cohorts, we only see some suggestive associations.

```{r}
manhattan(subset(single.snp, SNP %in% geno.SNPs2))

```

We got the same hits when considering SNPs genotyped in Epicuro and Obesit3 (biggest case cohort).

## Stratification by cohort

```{r, cache = TRUE}
obe1.snp <- getGWASTab(paste0("../results/associations/GWAS/", version, "/Obesity.GWAS.Obesity1.locuszoom.input.gz"))
obe2.snp <- getGWASTab(paste0("../results/associations/GWAS/", version, "/Obesity.GWAS.Obesity2.locuszoom.input.gz"))
obe3.snp <- getGWASTab(paste0("../results/associations/GWAS/", version, "/Obesity.GWAS.Obesity3.locuszoom.input.gz"))
```

```{r, cache = TRUE}
sigSNPs <- subset(single.snp, P < 5e-8)$SNP
info.list <- lapply(c("Epicuro", paste0("Obesity", 1:3)), 
                    getSNPimputationStatus, selSNPs = sigSNPs)
info.df <- Reduce(rbind, info.list)
table(info.df$Genotyped, info.df$Cohort)
```

Most of the significant SNPs were imputed. Nonetheless, the proportion of genotyped SNPs is higher in Epicuro, possibly due to its large array size.

```{r}
info.df %>%
    ggplot(aes(x = Cohort, y = Rsq, color = Cohort)) +
    geom_boxplot() +
    theme_bw()
info.df %>%
    group_by(Cohort) %>%
    summarize(median_Rsq = median(Rsq))
pairwise.wilcox.test(info.df$Rsq, info.df$Cohort)
```

There are no differences in imputation quality between the cohorts for variants with genome-wide significance.


```{r}
info.df %>%
    ggplot(aes(x = Cohort, y = as.numeric(EmpRsq), color = Cohort)) +
    geom_boxplot() +
    theme_bw()
info.df %>%
    group_by(Cohort) %>%
    summarize(median_LooRsq = median(as.numeric(EmpRsq), na.rm = TRUE))
pairwise.wilcox.test(as.numeric(info.df$EmpRsq ), info.df$Cohort)
```

When comparing the empirical Rsq, Epicuro has much worse values than the other cohorts.


```{r}
sig.allcohort <- single.snp %>%
    filter(SNP %in% sigSNPs) %>%
    mutate(Cohort = "Combined") %>%
    rbind(filter(obe1.snp, SNP %in% sigSNPs) %>% mutate(Cohort = "Obesity1")) %>%
    rbind(filter(obe2.snp, SNP %in% sigSNPs) %>% mutate(Cohort = "Obesity2")) %>%
    rbind(filter(obe3.snp, SNP %in% sigSNPs) %>% mutate(Cohort = "Obesity3")) 


sig.p <- sig.allcohort %>%
    dplyr::select(SNP, CHR, BP, P, Cohort) %>%
    mutate(P = -log10(P)) %>%
    spread(Cohort, P) 

sig.p.tmp <- sig.p %>%
    mutate(P.Combined = Combined,
           P.Obesity1 = Obesity1,
           P.Obesity2 = Obesity2,
           P.Obesity3 = Obesity3) %>%
    dplyr::select(SNP, starts_with("P."))

info.tmp <- cbind(info.df %>%
                      dplyr::select(SNP, Rsq, Cohort) %>%
                      mutate(Cohort = paste0("Rsq.", Cohort)) %>%
                      spread(Cohort, Rsq),  
                  info.df %>%
                      dplyr::select(SNP, EmpRsq, Cohort) %>%
                      mutate(Cohort = paste0("EmpRsq", Cohort)) %>%
                      spread(Cohort, EmpRsq))
    
sig.beta <- sig.allcohort %>%
    dplyr::select(SNP, CHR, BP, BETA, Cohort) %>%
    spread(Cohort, BETA) %>%
    left_join(sig.p.tmp, by = "SNP")  %>%
    left_join(dplyr::select(info.df, SNP, Rsq, Cohort) %>%
                  mutate(Cohort = paste0("Rsq.", Cohort)) %>%
                  spread(Cohort, Rsq), 
              by = "SNP")



beta.l <- lapply(paste0("Obesity", 1:3), function(x){
    sig.beta %>%
        ggplot(aes_string(x = "Combined", y = x)) +
        geom_point(alpha = 0.3) +
        scale_x_log10() +
        scale_y_log10() +
        geom_abline(slope = 1) +
        geom_hline(yintercept = 1) +
        geom_vline(xintercept = 1) +
        theme_bw() +
        ggtitle(x)
})
plot_grid(plotlist = beta.l, ncol = 3)

sig.p <- sig.allcohort %>%
    dplyr::select(SNP, CHR, BP, P, Cohort) %>%
    mutate(P = -log10(P)) %>%
    spread(Cohort, P) 

p.l <- lapply(paste0("Obesity", 1:3), function(x){
    ggplot(sig.p, aes_string(x = "Combined", y = x)) +
        geom_point(alpha = 0.3) +
        geom_abline(slope = 1) +
        theme_bw() +
        scale_x_continuous(limits = c(0, 43)) +
        scale_y_continuous(limits = c(0, 43)) +
        ggtitle(x)
})
plot_grid(plotlist = p.l, ncol = 3)
```

In general, the ORs are consistent between the full model and the subcohorts. Obesity1 presents the higher discrepancies, with a high number of variants with no association or inverse association in the subcohort with respect to the full model. On the other hand, Obesity3 had an almost perfect correlation with the full model estimate, possibly because is the biggest cohort and have a greater influence in the results. P-values are much smaller in the subcohorts and can only be compared in Obesity3, the biggest subcohort.

```{r}
sig.beta <- sig.beta %>% 
    mutate(direction = ifelse(Obesity1 > 1 & Obesity2 > 1 & Obesity3 > 1, "Positive",
                              ifelse(Obesity1 < 1 & Obesity2 < 1 & Obesity3 < 1, "Negative",
                                     "Incoherent")),
           coherence = ifelse(!is.na(direction) & direction != "Incoherent", "Coherent", "Incoherent"))


beta2.l <- lapply(paste0("Obesity", 1:3), function(x){
    sig.beta %>%
        ggplot(aes_string(x = "Combined", y = x)) +
        geom_point(alpha = 0.3) +
        scale_x_log10() +
        scale_y_log10() +
        geom_abline(slope = 1) +
        geom_hline(yintercept = 1) +
        geom_vline(xintercept = 1) +
        theme_bw() +
        facet_grid(~ coherence) +
        ggtitle(x)
})
plot_grid(plotlist = beta2.l, ncol = 3)


diff.l <- lapply(paste0("Obesity", 1:3), function(x){
     sig.beta %>%
        mutate(Pdiff = P.Combined - !!sym(paste0("P.", x)),
             Betadiff = Combined / !!sym(x)) %>%
     ggplot(aes(x = P.Combined, y = Betadiff)) +
        geom_point(alpha = 0.3) +
        theme_bw() +
        geom_hline(yintercept = 1) +
        scale_y_log10(name = "OR full / OR subcohort") +
        ggtitle(x) +
        facet_grid(direction ~ .)
})
plot_grid(plotlist = diff.l, ncol = 3)


diffcor.l <- lapply(paste0("Obesity", 1:3), function(x){
     sig.beta %>%
        mutate(Pdiff = P.Combined - !!sym(paste0("P.", x)),
               Betadiff = Combined / !!sym(x),
               Betadiff = abs(log(Betadiff))) %>%
    cor.test(~ Pdiff +  Betadiff, data =  .)
        
})
diffcor.l


```


Finally, we observed a small correlation between the differences in P-value between full cohort and the small cohorts and the difference in beta estimate between the full cohort and the main cohort. This result suggest that differences in significance are more related to lower power than to a different association. 

```{r}
diffP.Rsq.l <- lapply(paste0("Obesity", 1:3), function(x){
     sig.beta %>%
        mutate(Pdiff = P.Combined - !!sym(paste0("P.", x)),
               Rsq = !!sym(paste0("Rsq.", x))) %>%
    ggplot(aes(x = Pdiff, y = Rsq)) +
        geom_point(alpha = 0.1) +
        geom_smooth(method='lm') +
        theme_bw() +
        ggtitle(x)
})
plot_grid(plotlist = diffP.Rsq.l, ncol = 3)

diffP.Rsqcor.l <- lapply(paste0("Obesity", 1:3), function(x){
     sig.beta %>%
       mutate(Pdiff = P.Combined - !!sym(paste0("P.", x)),
               Rsq = !!sym(paste0("Rsq.", x))) %>%
     cor.test(~ Pdiff +  Rsq, data =  .)
        
})
diffP.Rsqcor.l
```

Only for Obesity3 we observed a correlation between the difference in P-value between the main and the small cohort model and the imputation quality. 


## Comparison with literature

We compared our results with the large meta-analysis of adult BMI (GIANT), children obesity (EGG consortium and SCOOP) and early-onset severe obesity (SCOOP). 

```{r}
giant <- read_table2("../data/bmi.giant-ukbb.meta-analysis.combined.23May2018.txt.gz")
egg <- read_table2("../data/CHILDHOOD_OBESITY.TRANS_ANCESTRAL.RESULTS.txt.gz")

scoop_obese <- read_table2("../data/SCOOP_STILTS_ldcorrected.gz")
scoop_extreme <- read_table2("../data/SCOOP_UKHLS_ldcorrected.gz")

giant$COORD <- paste(giant$CHR, giant$POS, sep = ":") 
egg$COORD <- paste(egg$CHR, egg$POS, sep = ":") 
scoop_obese$COORD <- paste(scoop_obese$CHR, scoop_obese$POS, sep = ":") 
scoop_extreme$COORD <- paste(scoop_extreme$CHR, scoop_extreme$POS, sep = ":") 
```

```{r, cache = TRUE}
single.snp$COORD <- paste(single.snp$CHR, single.snp$BP, sep = ":") 
sigCoords <- subset(single.snp, P < 5e-8)$COORD
sigCoordsGR <- GRanges(sigCoords)
sigCoords.filt <- sigCoords[-unique(from(findOverlaps(sigCoordsGR, LCRGR)))]
```

`r round(mean(sigCoords.filt %in% giant$COORD)*100, 2)`% of our significant SNPs were also reported in GIANT, `r round(mean(sigCoords.filt %in% egg$COORD)*100, 2)`% were also reported in EGG, while `r round(mean(sigCoords.filt %in% scoop_obese$COORD)*100, 2)`% SCOOP obesity and `r round(mean(sigCoords.filt %in% scoop_extreme$COORD)*100, 2)`% SCOOP extreme obesity. Then, we compared the estimates between our cohort and these meta-analysis. For EGG, we used the beta from the European population.

```{r, cache = TRUE}
loci.lit <- single.snp %>%
  mutate(BETA_OBE = BETA,
         P_OBE = P) %>%
  dplyr::select(COORD, ALLELE, BETA_OBE, P_OBE, SNP) %>%
  left_join(giant %>%
              mutate(BETA_GIANT = BETA,
                     P_GIANT = P) %>%
              dplyr::select(COORD, Tested_Allele, BETA_GIANT, P_GIANT),
            by = "COORD") %>%
    left_join(egg %>%
              mutate(BETA_EGG = as.numeric(EUR_BETA),
                     P_EGG = as.numeric(EUR_P)) %>%
              dplyr::select(COORD, EA, BETA_EGG, P_EGG),
            by = "COORD") %>%
    left_join(scoop_obese %>%
              mutate(BETA_SC_OB = OR,
                     P_SC_OB = P,
                     EA_SC_OB = EA) %>%
                dplyr::select(COORD, EA_SC_OB, BETA_SC_OB, P_SC_OB),
            by = "COORD") %>%
      left_join(scoop_extreme %>%
              mutate(BETA_SC_EXT = OR,
                     P_SC_EXT = P,
                     EA_SC_EXT = EA) %>%
              dplyr::select(COORD, EA_SC_EXT, BETA_SC_EXT, P_SC_EXT),
            by = "COORD")  %>%
  filter(P_OBE < 5e-8 | P_GIANT < 5e-8 | P_EGG < 5e-8 | P_SC_OB < 5e-8 | P_SC_EXT < 5e-8 | (SNP %in% geno.SNPs & P_OBE < 1e-5)) %>%
  mutate(BETA_EGG_mod = ifelse(toupper(EA) == ALLELE, BETA_EGG, -BETA_EGG),
         BETA_GIANT_mod = ifelse(Tested_Allele == ALLELE, BETA_GIANT, -BETA_GIANT),
         BETA_SC_OB_mod = ifelse(toupper(EA_SC_OB) == ALLELE, BETA_SC_OB, 1/BETA_SC_OB),
         BETA_SC_EXT_mod = ifelse(toupper(EA_SC_EXT) == ALLELE, BETA_SC_EXT, 1/BETA_SC_EXT),
         class = ifelse(P_OBE < 5e-8, ifelse(P_GIANT < 5e-8 | P_EGG < 5e-8 | P_SC_OB < 5e-8 | P_SC_EXT < 5e-8 , "Common", "OBE imputed"), ifelse(SNP %in% geno.SNPs & P_OBE < 1e-5, "OBE genotyped", "Literature")))
single.snp$COORD <- NULL
```

```{r}
p_egg <- ggplot(loci.lit, aes(x = -log10(P_OBE), y = -log10(P_EGG), color = class)) + 
  geom_point() +
  geom_hline(yintercept = 7.3) +
  geom_vline(xintercept = 7.3) +
  theme_bw() +
  scale_x_continuous("-log10 P Obesity") +
  ggtitle("Children Obesity EGG")

p_giant <- ggplot(loci.lit, aes(x = -log10(P_OBE), y = -log10(P_GIANT), color = class)) + 
  geom_point() +
  geom_hline(yintercept = 7.3) +
  geom_vline(xintercept = 7.3) +
  theme_bw() +
  scale_x_continuous("-log10 P Obesity") +
  ggtitle("BMI Adults")

 p_ob <- ggplot(loci.lit, aes(x = -log10(P_OBE), y = -log10(P_SC_OB), color = class)) + 
  geom_point() +
  geom_hline(yintercept = 7.3) +
  geom_vline(xintercept = 7.3) +
  theme_bw() +
  scale_x_continuous("-log10 P Obesity") +
  ggtitle("Children Obesity SCOOP")

  p_ext <- ggplot(loci.lit, aes(x = -log10(P_OBE), y = -log10(P_SC_EXT), color = class)) + 
  geom_point() +
  geom_hline(yintercept = 7.3) +
  geom_vline(xintercept = 7.3) +
  theme_bw() +
  scale_x_continuous("-log10 P Obesity") +
  ggtitle("Extreme Children Obesity SCOOP")
plot_grid(p_ext, p_ob, p_egg, p_giant, ncol = 2)
```

There is no correlation between the p-values in our GWAS and the p-values from the literature. Thus, the hits identified in our cohort have not been previously identified, while the hits identified in GIANT, EGG and SCOOP have not been found in our cohort. 

```{r}
beta_giant <- loci.lit %>%
  filter((P_OBE < 5e-8 | P_GIANT < 5e-8 | SNP %in% geno.SNPs) & !is.na(class)) %>%
  ggplot(aes(x = BETA_OBE, y = BETA_GIANT_mod, color = class)) + 
  geom_point() +
  scale_x_log10("Obesity OR") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 1) +
  theme_bw()  +
  facet_grid(~ class, scales = "free") +
  ggtitle("BMI Adults")

beta_egg <-  loci.lit %>%
  filter((P_OBE < 5e-8 | P_EGG < 5e-8 | SNP %in% geno.SNPs) & !is.na(class)) %>%
  ggplot(aes(x = BETA_OBE, y = exp(BETA_EGG_mod), color = class)) + 
  geom_point() +
  scale_x_log10("Obesity OR") +
  scale_y_log10() +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) +
  geom_abline(slope = 1) +
  theme_bw() +
  facet_grid(~ class, scales = "free") +
  ggtitle("Children Obesity EGG")

beta_ob <-  loci.lit %>%
  filter((P_OBE < 5e-8 | P_SC_OB < 5e-8  | SNP %in% geno.SNPs) & !is.na(class)) %>%
  ggplot(aes(x = BETA_OBE, y = BETA_SC_OB_mod, color = class)) + 
  geom_point() +
  scale_x_log10("Obesity OR") +
  scale_y_log10() +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) +
  geom_abline(slope = 1) +
  theme_bw() +
  facet_grid(~ class, scales = "free") +
  ggtitle("Children Obesity SCOOP")

beta_ext <-  loci.lit %>%
  filter((P_OBE < 5e-8 | P_SC_EXT < 5e-8  | SNP %in% geno.SNPs) & !is.na(class)) %>%
  ggplot(aes(x = BETA_OBE, y = BETA_SC_EXT_mod, color = class)) + 
  geom_point() +
  scale_x_log10("Obesity OR") +
  scale_y_log10() +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) +
  geom_abline(slope = 1) +
  theme_bw() +
  facet_grid(~ class, scales = "free") +
  ggtitle("Extreme Children Obesity SCOOP")


plot_grid(beta_ext, beta_ob, beta_egg, beta_giant, nrow = 2)


```



Only one of the hits in our genotyped subset had signicant ORs in any of the datasets. For the other hits, the OR in the literature was close to 0. 

```{r}
subset(loci.lit, class == "OBE genotyped") %>% data.frame()
```

```{r}
cohorts <- c("SC_EXT", "SC_OB", "EGG", "GIANT")
names(cohorts) <- cohorts
lapply(cohorts, function(x) {
  if ( x == "EGG"){
      cor.test(formula(~ exp(BETA_EGG_mod) + BETA_OBE), loci.lit, subset = loci.lit$class == "Literature")
  } else{
      cor.test(formula(paste0("~ BETA_", x, "_mod + BETA_OBE")), loci.lit, subset = loci.lit$class == "Literature")
  }
} )

```
Despite the literature hits had big p-values in our cohort, the inferred beta values are moderately correlated (r around 0.25). Interestingly, we observed that for children obesity in SCOOP, ORs in our GWAS were more extreme than those originally reported, although none of these hits was significant in our GWAS.

```{r}
lapply(cohorts, function(x) {
    if ( x == "EGG"){
      cor.test(formula(~ exp(BETA_EGG_mod) + BETA_OBE), loci.lit, subset = loci.lit$class == "OBE imputed")
    } else{
        cor.test(formula(paste0("~ BETA_", x, "_mod + BETA_OBE")), loci.lit, subset = loci.lit$class == "OBE imputed")

    }

} )
```

On the other hand, we observed very little correlation between the beta values of the hits detected in our GWAS and in the literature. The main exception was with hits from Extreme children obesity, where we achieved a high correlation. Unexpectedly, we observed a strong but inverse correlation between our GWAS beta-values and beta values from EGG.

```{r}
p_lit <- loci.lit %>%
  filter(class == "Literature") %>%
  ggplot(aes(x = -log10(P_EGG), y = -log10(P_SC_OB))) + 
  geom_point() +
  scale_x_continuous("EGG -log10 P") +
  scale_y_continuous("SCOOP -log10 P") +
  geom_hline(yintercept = 7.3) +
  geom_vline(xintercept = 7.3) +
  theme_bw()

beta_lit <- loci.lit %>%
  filter(class == "Literature") %>%
  ggplot(aes(x = exp(BETA_EGG_mod), y = BETA_SC_OB_mod)) + 
   scale_x_log10("EGG OR") +
  scale_y_log10("SCOOP OR") +
  geom_point() +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) +
  geom_abline(slope = 1) +
  theme_bw()
plot_grid(p_lit, beta_lit, nrow = 1)
```

Finally, we compared the p-values and the betas from the two GWAS of children obesity. In this case, the p-values are highly correlated, although they are smaller in EGG, the cohort with the biggest sample size. Regarding the ORs, there are also more extreme in EGG, although they are mostly coherent. 
```{r}
rm(loci.lit, giant, egg, scoop_obese, scoop_extreme)
gc()
```

## Adjustment for genetic components

```{r, cache = TRUE}
raw.snp <- getGWASTab(paste0("../results/associations/GWAS/", version, "/Obesity.GWAS.plink.raw.locuszoom.input.gz"))
```

```{r}
sig.raw <- single.snp %>%
    filter(P < 5e-8) %>%
    mutate(Model = "Full") %>%
    rbind(filter(raw.snp, P < 5e-8) %>% mutate(Model = "Crude")) 

raw.beta <- sig.raw %>%
  dplyr::select(SNP, CHR, BP, BETA, Model) %>%
  spread(Model, BETA) %>%
  ggplot(aes(x = Full, y = Crude)) +
        geom_point(alpha = 0.3) +
        scale_x_log10() +
        scale_y_log10() +
        geom_abline(slope = 1) +
        geom_hline(yintercept = 1) +
        geom_vline(xintercept = 1) +
        theme_bw() +
        ggtitle("OR comparison")

raw.p  <- sig.raw %>%
  dplyr::select(SNP, CHR, BP, P, Model) %>%
    mutate(P = -log10(P)) %>%
    spread(Model, P) %>% 
  ggplot(aes(x = Full, y = Crude)) +
        geom_point(alpha = 0.3) +
        geom_abline(slope = 1) +
        theme_bw() +
        scale_x_continuous(limits = c(0, 43)) +
        scale_y_continuous(limits = c(0, 43)) +
        ggtitle("-log10 P comparison")

plot_grid(raw.beta, raw.p, ncol = 2)

sig.raw %>%
  dplyr::select(SNP, CHR, BP, BETA, Model) %>%
  spread(Model, BETA) %>%
  cor.test(~ Full + Crude, data = .)
```

The coefficients are almost identical in the crude or the adjusted model. The p-values are correlated between both models but with p-values from crude model slightly more significant. There are no hits that are very significant only in one of the models. 


## Sensitivity

We run the GWAS after excluding the individuals with potential pathogenic SNVs or CNVs. 

```{r, cache = TRUE}
path.snp <- getGWASTab(paste0("../results/associations/GWAS/", version,  "/Obesity.GWAS.plink.non_patho.locuszoom.input.gz"))

```
```{r}
sig.patho <- single.snp %>%
    filter(SNP %in% sigSNPs) %>%
    mutate(Cohort = "Full") %>%
    rbind(filter(path.snp, SNP %in% sigSNPs) %>% mutate(Cohort = "Sensitivity")) 

sense.beta <- sig.patho %>%
  dplyr::select(SNP, CHR, BP, BETA, Cohort) %>%
  spread(Cohort, BETA) %>%
  ggplot(aes(x = Full, y = Sensitivity)) +
        geom_point(alpha = 0.3) +
        scale_x_log10() +
        scale_y_log10() +
        geom_abline(slope = 1) +
        geom_hline(yintercept = 1) +
        geom_vline(xintercept = 1) +
        theme_bw() +
        ggtitle("OR comparison")

sense.p  <- sig.patho %>%
    dplyr::select(SNP, CHR, BP, P, Cohort) %>%
    mutate(P = -log10(P)) %>%
    spread(Cohort, P) %>% 
  ggplot(aes(x = Full, y = Sensitivity)) +
        geom_point(alpha = 0.3) +
        geom_abline(slope = 1) +
        theme_bw() +
        scale_x_continuous(limits = c(0, 43)) +
        scale_y_continuous(limits = c(0, 43)) +
        ggtitle("-log10 P comparison")

plot_grid(sense.beta, sense.p, ncol = 2)

sig.patho %>%
  dplyr::select(SNP, CHR, BP, BETA, Cohort) %>%
  spread(Cohort, BETA) %>%
  cor.test(~ Full + Sensitivity, data = .)
```

The coefficients have not changed after removing individuals with candidate pathogenic variants. As expected, the p-values are bigger due to the reduced sample size. 

Therefore, the individuals with pathogenic variants share the common variants that increase risk to obesity with the individuals without pathogenic variants. 

# Inversions

We associated the chromosomal inversions from scoreInvHap with extreme obesity. We adjusted all the models by the first 10 principal components of genetic variance. 

```{r}
load(paste0("../results/associations/inversions/", version, "/Obesity.inversions.assoc.Rdata"))
inv_assocs
```
Only inv16_009 was associated with extreme obesity.

```{r}
attr(inv_assocs, "tables")$inv16_009
```

Inverted allele is more frequent in case samples. The results support a dominant model, where having at least one inverted allele have 1.39 more chances of having obesity. 

## Inv16 haplotypes

Due to the polymorphic nature of inv16_009, we aimed to further explore the association of inv16 with the phenotypes. Current implementation of scoreInvHap does not discriminate between inv16 haplotypes, therefore, we relied on multidimensional scaling (MDS), the method that used invClust for genotyping.

```{r}
load(paste0("../results/associations/inversions/", version, "/ObesityscoreInvHapClassDF.Rdata"))
mds.df <- mds %>%
  data.frame() %>%
  mutate(id = rownames(.)) %>%
  left_join(mutate(scClassDF, id = rownames(.)), by = "id") %>%
  tibble()

mds.df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = inv16_009_haplos)) +
  geom_point() +
  theme_bw()
```

We defined 6 haplotypes in the data. Based in previous results, we defined two of the haplotypes as standard and the remaining as inverted.


```{r}
mds.df %>%
  ggplot(aes(x = MDS1, y = MDS2, color = inv16_009)) +
  geom_point() +
  theme_bw()
table(scoreInvHap = mds.df$inv16_009, MDS = mds.df$inv16_009_haplos)
```

We observed some discrepancies between the scoreInvHap classification and the MDS classification. 

```{r}
mds.df$inv16_mds <- ifelse(mds.df$inv16_009_haplos == "II", "II",
                           ifelse(mds.df$inv16_009_haplos %in% c("NaI", "NbI"), "NI", "NN"))
table(scoreInvHap = mds.df$inv16_009, MDS = mds.df$inv16_009_haplos)
tab <- table(scoreInvHap = mds.df$inv16_009, MDS = mds.df$inv16_mds)
tab
```

`r round((sum(tab) - sum(diag(tab)))/sum(tab)*100, 2)`% of samples have a discordant inversion status between the MDS and scoreInvHap classifications.

We run the association analysis using the genotypes defined with the MDS. We considered each of the alleles independently. Of note, the allele I correspond to the same analysis using scoreInvHap but with a different algorithm to classify the samples:

```{r}
inv16_assocs
tabs <- attr(inv16_assocs, "tables")
tabs
```

In this setting, A means alternate allele and R the reference allele. For instance, in inv16_I, A/A means a II individual while RR is an NaNa, NaNb o NbNb individual. The association between the I allele and the extreme obesity has a reduced magnitude and it is only associated with the dominant and almost significant with the additive, suggesting that the I allele increases the risk of having extreme obesity. 

# Genetic scores

We obtained polygenic risk scores from a model of BMI obtained from GWAS catalog (PGS000027).

```{r}
iniQCfam <- dir(path = "../results/preImputation/plink/", pattern = "fam",
  full.names = TRUE, recursive = TRUE)
names(iniQCfam) <- gsub(".fam", "", basename(iniQCfam))

famObj <- lapply(iniQCfam, read.delim, header = FALSE, sep = " ")
famDf <- Reduce(rbind, famObj) %>%
  mutate(Dataset = rep(names(iniQCfam), sapply(famObj, nrow)),
         IID = V2) %>%
  dplyr::select(IID, Dataset)

pcs <- read.table("../results/postImputation/stats/2021-05-25/mergedDataset.filtered.eigenvec")
colnames(pcs) <- c("FID", "IID", paste0("PC", 1:10))

pgs <- read.table(paste0("../results/associations/PRS/", version, "/obesity.profile"), 
                  header = TRUE) %>%
    mutate(STATUS = ifelse(grepl("SB", IID), "Control", "Case"),
           STATUS = factor(STATUS, levels = c("Control", "Case")),
           PGS = (SCORESUM - mean(SCORESUM))/sd(SCORESUM), 
           PGS_quantile = cut(PGS, 
                              breaks = quantile(PGS, seq(0, 1, 0.1)),
                              labels = paste0("Q", 1:10),
                              include.lowest = TRUE),
           PGS_carrier = ifelse(PGS_quantile == "Q10", "Carrier", "Reference"),
           PGS_carrier = factor(PGS_carrier, levels = c("Reference", "Carrier"))) %>%
    left_join(dplyr::select(pcs, -FID), by = "IID") %>%
    left_join(famDf, by = "IID") %>%
    as_tibble()

```
```{r}
pgs %>%
    ggplot(aes(x = PGS, color = STATUS)) +
    geom_density() +
    theme_bw() +
    geom_vline(xintercept = 0)

pgs %>% 
    glm(STATUS ~ PGS, ., family = "binomial") %>%
    summary()

pgs %>% 
    dplyr::select(starts_with("PC"), STATUS, PGS) %>%
    glm(STATUS ~ PGS + ., ., family = "binomial") %>%
    summary()


pgs %>% 
    dplyr::select(starts_with("PC"), STATUS, PGS) %>%
    lm(PGS ~ STATUS + ., .) %>%
    summary()
```

Case samples have higher polygenic risk scores than control samples. This difference is significant even after adjusting for genetic principal components. 



```{r}
pgs_Qtab <- pgs %>%
    mutate(PGS_cat = ifelse(PGS_quantile == "Q1", "Bottom quantile", 
                            ifelse(PGS_quantile == "Q10", "Top quantile", 
                                   "Quantiles 2-9")),
           PGS_cat = factor(PGS_cat, 
                            levels = c("Bottom quantile", "Quantiles 2-9", "Top quantile"))) %>%
    group_by(STATUS, PGS_cat) %>%
    summarize(n = n()) %>%
    group_by(PGS_cat) %>%
    mutate(prop = n/sum(n))
pgs_Qtab %>%
    ggplot(aes(x = PGS_cat, y = prop, fill = PGS_cat)) +
    geom_bar(stat = "identity", colour = "black") +
    facet_grid(~ STATUS) +
    xlab("Disease Status") +
    ylab("Percent") +
    scale_fill_manual(name = "Polygenic score", values = c("#EEEEEE", "#999999", "#444444"))
```

The proportion of individuals with extreme obesity is much higher in the top quantile than in the bottom quantile.

```{r}
pgs %>% 
    dplyr::select(starts_with("PC"), STATUS, PGS_carrier) %>%
    glm(STATUS ~ ., ., family = "binomial") %>%
    summary()

pgs %>% 
    dplyr::select(starts_with("PC"), STATUS, PGS_carrier) %>%
    glm(STATUS ~ ., ., family = "binomial") %>%
    logistic.display()
```

Individuals in the top decile of the PRS have an OR of 3.09 of being extreme obese.

## Influence of rare variants

We explored whether rare variants (CNVs or SNVs) identified in these individuals had an effect on the polygenic score.

```{r}
sampVariants <- read_excel("../data/Obesity.SamplesSummary.xlsx") %>%
  filter(!is.na(Plate))

pathVars <- sampVariants %>%
  filter(!is.na(`RSVs Over-Patho`)  | !is.na(`RSVs Patho`))

cnvs <- sampVariants %>%
  filter(!is.na(rCNVs) | !is.na(`comCNVs`))


pgs <- pgs %>%
  mutate(STATUS2 = ifelse(STATUS == "Case" & IID %in% pathVars$Patient, "Case_pathogenic",
                          ifelse(STATUS == "Case" & IID %in% cnvs$Patient, "Case_cnv", 
                                 ifelse(STATUS == "Case", "Case_novariants", "Control"))),
         STATUS2 = factor(STATUS2, levels = c("Case_novariants", "Case_cnv", "Case_pathogenic", "Control")),
        STATUS3 = ifelse(STATUS2 == "Case_cnv", "Case_pathogenic", as.character(STATUS2))) %>%
  distinct()
  

```

We classified the case individuals in three groups:

- pathogenic: individuals with pathogenic variants and potentially penetrant variants
- cnvs: individuals with CNVs that can cause the disease
- novariants: individuals without any candidate genetic variant.

```{r}
pgs %>%
    ggplot(aes(x = PGS, color = STATUS2)) +
    geom_density() +
    theme_bw() +
    geom_vline(xintercept = 0)

pgs %>% 
    lm(PGS ~ STATUS2, .) %>%
    summary()

pgs %>% 
    dplyr::select(starts_with("PC"), STATUS2, PGS) %>%
    lm(PGS ~ ., .) %>%
    summary()

pgs %>% 
    dplyr::select(starts_with("PC"), STATUS3, PGS) %>%
    lm(PGS ~ ., .) %>%
    summary()
```

Although the PGS seems slightly higher for cases with pathogenic variants, there are so few samples and the associations are not statistically significant.



## Stratification by cohort

We examined whether we could find differences between the cohorts, which can be caused by technical differences during implementation.

```{r}
pgs %>%
    ggplot(aes(x = PGS, color = Dataset)) +
    geom_density() +
    theme_bw() +
    geom_vline(xintercept = 0)

pgs %>% 
    dplyr::select(starts_with("PC"), Dataset, PGS) %>%
    lm(PGS ~ ., .) %>%
    summary()

pgs %>% 
    dplyr::select(starts_with("PC"), Dataset, PGS) %>%
    mutate(Dataset = factor(Dataset, levels = c("Obesity2", "Obesity1", "Obesity3", "Epicuro"))) %>%
    lm(PGS ~ ., .) %>%
    summary()
```

Individuals from Obesity2 have a lower PGS than individuals from Obesity1 or Obesity3.